Modeling the Latent Space of Symbolic String Quartet Music

### 2021-06-01 papers/final/alex/paper.tex
separate final paper to individual
alex paper updates

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Musical Deep Learning}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Nhut Minh Phan}
\email{phan92@uw.edu}
\affiliation{%
  \institution{University of Washington}
  \city{Bothell}
  \state{WA}
  \country{USA}
}

@@

\begin{abstract}
  This paper presents a generative model of multi-instrument symbolic
  music.
\end{abstract}

\keywords{deep learning, neural networks, music}

@@

Machine learning models of music have interesting applications in
music information retrieval and creative tools for musical artists and
educators. Generative models can create accompaniments for music,
transfer styles between clips of music, and even generate entirely new
music. Music is challenging to model because it exhibits a complex
hierarchy of recurring patterns and long-range temporal dependencies,
and because both musical scores and performances have multiple
possible digital representations.

Depending on the task, machine learning models of music may be trained
on the audio signal itself, either in a time domain or a frequency
domain representation, or they may be trained on a digital symbolic
representation of music, the most common of which is MIDI (Musical
Instrument Digital Interface) notation. MIDI is an encoding of music
as streams of bytes in one or more tracks or channels, each
representing a sequence of 128 possible pitch values (where 0 is the
lowest and 127 is the highest), along with timing, pressure and
instrument identifier values. A music transcription model may
transcribe an audio signal as a MIDI score, which can easily be
converted into other symbolic representations such as sheet music for
human performers to read from, while a synthesizer model can convert
MIDI representations into audio signals. A generative music model can
be trained either to generate raw audio, or to produce a symbolic
score that must be played by a synthesizer or by humans to produce an
audio music performance. This project focuses on the latter type of
modeling: generative modeling of symbolic (MIDI) music to compose
original musical scores.

@@

The state of the art in music generation has a long way to go before
it can consistently generate music scores or performances that would
be enjoyable and popular for humans to listen to, but a number of
recent research projects have shown promising progress in this area.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which we plan to use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}. Including this second dataset may
help to generalize the modeling approach beyond European classical
music to include other popular genres and associated instruments.

@@

The result of this preprocessing is that each training example will be
a 3D tensor of shape (tracks x ticks x pitches) and stacking the
training examples will produce a 4D tensor.

While we plan to model music with multiple instrument tracks, we
anticipate the need to model only a fixed selection of instrument
parts, similar to how MusicVAE models three-part (drum, bass and
melody) \cite{roberts_hierarchical_2018} and MuseGAN models five-part
(drum, bass, guitar, string, piano) arrangements.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_bar.png}
  \caption{Pianoroll visualization of the first measure of
    Beethoven's Serioso String Quartet}
  \label{pianoroll}
  \Description{A pianoroll representation of cello, viola and violin
    parts from a Beethoven piece.}
\end{figure}

@@

>>>>

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Modeling the Latent Space of Symbolic String Quartet Music}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

@@

\begin{abstract}
  This paper presents the results of a project to apply a recurrent
  variational autoencoder to the task of modeling a latent space for
  generating pieces of multi-instrument symbolic classical music
  through interpolation.
\end{abstract}

\keywords{deep learning, recurrent neural networks, sequential models,
  music modeling, latent space modeling, generative modeling}

@@

Machine learning models of music have interesting applications in
music information retrieval and creative tools for musical artists and
educators. Generative models can create accompaniments for music,
blend and transfer styles between clips of music, and even generate
entirely new music. Music is challenging to model because it is
inherently high-dimensional, exhibiting a complex hierarchy of
recurring patterns and long-range temporal dependencies, and because
musical scores have multiple possible digital representations with
distinct advantages and disadvantages.

Depending on the task, machine learning models of music may be trained
on the audio signal of a musical performance, either in a time domain
or a frequency domain representation, or they may be trained on a
digital symbolic representation of music, the most common of which is
MIDI (Musical Instrument Digital Interface) notation. MIDI is an
encoding of music as streams of bytes in one or more tracks or
channels, each representing a sequence of 128 possible pitch values
(where 0 is the lowest pitch and 127 is the highest), along with
timing, pressure and instrument identifier values. A generative
symbolic music model can produce a symbolic score in MIDI format,
which must be played by synthesizer software or by humans to produce
an audio music performance. This project focuses on generative
modeling of symbolic (MIDI) music to compose original musical scores
by blending input scores through continuous latent space interpolation.

@@

The state of the art in music generation still has a long way to go
before it can consistently generate music scores or performances that
would be enjoyable and popular for humans to listen to, but for this
reason it is an area of significant opportunity, where a number of
recent research projects have shown promising progress.

@@

The dataset used for this research project is MusicNet, which is a
collection of 330 freely licensed European classical music recordings
with aligned MIDI scores \cite{thickstun2017learning}.  The model is
trained on 36 string quartets (four-part arrangements with two
violins, one viola and one cello) by composers Bach, Beethoven,
Dvorak, Haydn, Mozart, and Ravel.

@@

The result of this preprocessing is that each training example will be
a 2D matrix of shape (time steps x instrument tracks) and stacking the
training examples will produce a 3D tensor. Information regarding which
track corresponds to which musical instrument is lost and therefore
must be maintained separately to reconstruct the MIDI representation.

As the model will produce one output per instrument track, it can
incorporate only a fixed selection of instrument parts, similar to how
MusicVAE models three-part (drum, bass and melody)
\cite{roberts_hierarchical_2018} and MuseGAN models five-part (drum,
bass, guitar, string, piano) arrangements. Therefore this model is
purpose-built for string quartets, which are the most common type of
arrangement found in the MusicNet dataset.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_bar.png}
  \caption{Pianoroll visualization of the first measure of
    Beethoven's Serioso String Quartet in F Minor}
  \label{pianoroll}
  \Description{A pianoroll representation of cello, viola and violin
    parts from a Beethoven piece.}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_bar_sheet.png}
  \caption{Sheet music of the first measure of Beethoven's Serioso
    String Quartet in F Minor}
  \label{sheet}
  \Description{A measure of sheet music cello, viola, and violin parts
    from Beethoven piece.}
\end{figure}

@@

The model is a recurrent variational autoencoder; both Long Short-Term
Memory (LSTM) and Gated Recurrent Unit (GRU) cell types were tested.
